### 2023-12-15 main.tex
''

The model studied in this paper is the {\em  multispecies stirring process with open boundaries}, i.e. the multispecies analogue of the exclusion process on a generic graph as introduced in \cite{zhou2021orthogonal} with additional boundary reservoirs as proposed in \cite{vanicat2017exact}. 
This particle system is defined on a generic connected and undirected graph. At each site 
the maximal number of allowed particles {\color{blue}(also called \textit{maximal occupancy})} is $\nu\in \mathbb{N}$. These particles can be of $N-1$ types (or species). Two sites do interact when there is an edge between them and this interaction consists in swapping the occupations of these two sites (for a precise definition of the process see Section \ref{sec1}). 
Furthermore, reservoirs are attached at each site of the graph, creating and removing particle. The process becomes integrable when the graph is a finite chain with two reservoirs attached to the first and the last sites respectively and the maximal number of allowed particles is fixed to be $\nu=1$,  see \cite{vanicat2017exact}. {\color{blue}As we will see, this choice of the boundary reservoirs is further motivated by the fact that they allow to determine an absorbing dual process, regardless of the geometry on which the process is defined and  the maximal number of particle hosted at each site ($\nu$ can any natural number)}.\\
{\color{blue}In this paper we concentrate our effort in studying the boundary driven symmetric multispecies model. In literature, also asymmetric multispecies  models have been introduced \cite{schutz2017kardar} and integrability results have been proved both with \cite{crampe2016integrable, finn2018matrix} and without boundary driving \cite{mallick1,mallick2}. Moreover, different transition mechanisms (beyond the diffusive one) have been recently studied \cite{chatterjee2023multi}. Finally, the presence of different types of particles have also been modelled via a "multi-layer" dynamics (see \cite{redig2022ergodic,floreani2022switching}). As far as we know, in this last context integrability does not hold.} 
%{\color{red}Explain the model here.... $N$-species (or $N-1$ plus empty?). What is the graph? say that it becomes integrable for the line }
%We refer to the model as the {\em  multispecies stirring process with open boundaries}.
%

@@

{\color{blue} Here $\mu(\bm{\tau})$ is the  non-equilibrium steady probability of having a configuration $\bm{\tau}$, that can be composed by particles and holes as well. However, from Corollary \ref{Corolollary-ABS_Corr-abstract}, we only have an explicit formula for the expectations of the indicator of particle configurations (indeed $a_{1},\ldots,a_{m}$ are different from $N$). Therefore, to write the probability mass function $\mu$ for occupations containing holes we have to use the conservation of probability to get rid of the indicators of these holes. Then, equation \eqref{implicit-mass-function} is derived (for details see below the proof of Corollary \ref{Corollary-NESS-ABS}).\\
 The expectation on the right-hand-side of \eqref{implicit-mass-function} contains only indicators of particles configurations: indeed in the first product  $\tau_{x}\neq N$ are considered, while in the second product we have $a_{k}\neq N$ for all $k\in\{1,\ldots,N\}$. As a consequence, we can map the coordinates $x_{1},\ldots,x_{m}$ and the types $a_{1},\ldots,a_{m}$ introduced in Corollary \ref{Corolollary-ABS_Corr-abstract} with, respectively, the coordinates $x$ (where the $\tau_{x}\neq N$) and $q_{1},\ldots,q_{b}$ and with the types $\tau_{x}\neq N$ and $a_{1},\ldots,a_{b}$. For this reason, by knowing the correlations \eqref{exactCorrelations}, the probability mass function $\mu(\bm{\tau})$ given in equation \eqref{implicit-mass-function} is obtained.}\\

@@

 	Here the right-hand-side corresponds to $b=0$ and average can be computed by Corollary \ref{Corolollary-ABS_Corr-abstract} with $m=2$, $x_{1}=1,\;x_{2}=2$ and $a_{1}=a_{2}=1$.

@@

 		\mu(1,N)=&\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=1\}}\mathbbm{1}_{\{Y_{2}=1\}}\right]=\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=1\}}\left(1-\sum_{a=1}^{N-1}\mathbbm{1}_{\{Y_{2}=a\}}\right)\right]\nonumber

@@

 	Here the first term on the right-hand-side corresponds to $b=0$ in \eqref{implicit-mass-function} and can be computed by Corollary \ref{Corolollary-ABS_Corr-abstract} with $m=1$, $x_{1}=1$ and $a_{1}=1$. The second term on the right-hand-side corresponds to $b=1$ in \eqref{implicit-mass-function} and can be computed by Corollary \ref{Corolollary-ABS_Corr-abstract} with $m=2$, $x_{1}=1,\;x_{2}=2$ and $a_{1}=1,\;a_{2}=a$ for all $a=1,\ldots N-1$. 

@@

 		\mu(N,N)=&\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=1\}}\mathbbm{1}_{\{Y_{2}=1\}}\right]=\mathbb{E}\left[\left(1-\sum_{a_{1}=1}^{N-1}\mathbbm{1}_{\{Y_{1}=_{1}a\}}\right)\left(1-\sum_{a_{2}=1}^{N-1}\mathbbm{1}_{\{Y_{2}=a_{2}\}}\right)\right]\nonumber

@@

 		1-\sum_{a_{1}=1}^{N-1}\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=a_{1}\}}\right]-\sum_{a_{2}=1}^{N-1}\mathbb{E}\left[\mathbbm{1}_{Y_{2}=a_{2}}\right]+\sum_{a_{1},a_{2}=1}^{N-1}\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=a_{1}\}}\mathbbm{1}_{\{Y_{2}=a_{2}\}}\right]\dt
 	Here the first term on the right-hand-side corresponds to $b=0$ and since $\bm{\tau}=(N,N)$ we obtain $1$. The second and the third term on the right-hand-side correspond to $b=1$. They can be computed by Corollary \ref{Corolollary-ABS_Corr-abstract} with $m=1$ and by fixing $x_{1}=1$, $a_{1}=1,\ldots,N-1$ and $x_{2}=2$, $a_{2}=1,\ldots,N-1$ respectively. The fourth term on the right-hand-side corresponds to $b=2$ and it can be computed by Corollary \ref{Corolollary-ABS_Corr-abstract} with $m=2$, $x_{1}=1,\,x_{2}=2$ and $a_{1},a_{2}=1,\ldots,N-1$. 

@@

			\langle \bm{\tau}|S_{1}^{-1}|\Psi^{'}\rangle=&\sum_{\bm{\sigma}\in \Omega^{'}}\Psi^{'}(\bm{\sigma})\prod_{x=1}^{L}\left(\delta_{\tau_{x},\sigma_{x}}+(-1)\delta_{\tau_{x},N}\sum_{a=1}^{N-1}\delta_{\sigma_x,a}\right)\nonumber\\

@@

		means that, {given $\bm{\tau}$ from the bra-vector $\langle \bm{\tau}|$ introduced before equation \eqref{Tensor-S1},} for all $b\in \{0,\ldots,L\}$ and for all $1\leq q_{1}<\ldots<q_{b}\leq L$ {at sites $q_{k}$ (for $k=1,\ldots,b$) we have a particle of type $a_{k}$, while at the other sites the occupation $\tau_{x}$ is left unchanged  ($x\notin\{q_{1},\ldots,q_{b}\}$). }  

@@

\section{Extensions beyond the integrable chain}
\label{sec5}
{\color{blue}In Section \ref{sectionIntegrabiliy} we have shown that, combining duality and integrability of the multispcies stirring process with $\nu=1$ and on the geometry of chain with two boundaries, the non-equilibrium steady state can be found in an explicit form. As soon as $\nu>1$ the integrability property is lost, but duality still holds. Therefore, in this section we first address the  question whether we can use duality to characterize the non-equilibrium steady state of the boundary driven multispecies stirring process on a chain with $\nu>1$ via absorption probabilities. For the sake of simplicity and  for the sake of clarity, we keep the geometry of a chain with two reservoirs attached to the end sites $1$ and $L$ of the chain. However, this question can be extended to the geometry of the graph $G$ introduced in Section \ref{sec1}.  \\
Inspired by the results for the thermalized version of the symmetric exclusion process reported in \cite{carinci2013duality}, in this section we also investigate its multispecies counterpart. More precisely, we try to answer these questions: \textit{can we define a thermalized version of the boundary driven multispecies stirring process? Does this thermalized process keep the absorbing duality properties as in the single species case?} Again, we restrict our analysis on the chain with two reservoirs at the end sites. \\ Finally we add a reaction mechanism of transition to the boundary driven multispeices stirring process. The aim is to answer the following question: \textit{Can we generalize to the case with $\nu>1$ with arbitrary number of species the boundary driven model introduced in \cite{casini2022uphill}?} Moreover, in \cite{casini2022uphill} the duality for this reaction-diffusion process was proved only for the equilibrium case and the result was achieved by direct computation (without using a $gl(3)$ symmetry). Thus, we also investigate if this last duality result can be extended to a boundary driven process by using an approach similar to the one of Section \ref{sectionDuality} and based on Lie algebras.\\ The section is organized as follows: }

@@

	\item \textit{Reaction-diffusion process} (Section~\ref{Subsection-RD}): {\color{blue} starting from the usual stirring dynamics (see Section \ref{sec1}) where particles and holes are exchanged in two nearest neighbour sites (now at a constant rate $\sigma_{11}$),} {\color{blue}two further reaction} dynamics are added. Particle can now change type {\color{blue} with two different mechanism: a \textit{pure mutation} (with rate $\Upsilon$) and a \textit{stirring-mutation} (at rate $\sigma_{12}$), i.e. a transition where particles perform an exchange of their occupancies at nearest neighbour sites but, at the same time, they also mutate their own species.} {\color{blue} This reaction-diffusion} model   is carefully constructed with the goal of having the average densities evolving as a closed system of $N-1$ coupled difference-differential equation, providing a generalization of the model studied in \cite{casini2022uphill}. {\color{blue} Exploiting the $gl(N)$ symmetry of this reaction-diffusion process, absorbing duality} for the boundary driven process is also proved.

@@

{\color{blue} We recall a general fact. Considering a generic $N-1$ species Markov process $(\bm{n}(t))_{t\geq 0}$ (where in the spirit of Section \ref{sec1}, $n_{A}^{x}$ denotes the number of particles or the number of holes at site $x$), we define the average density occupation of species $a\in\{1,\ldots,N-1\}$ at site $x$ and at time $t$ as

>>>>

The model studied in this paper is the {\em  multispecies stirring process with open boundaries}, i.e. the multispecies analogue of the exclusion process on a generic graph as introduced in \cite{zhou2021orthogonal} with additional boundary reservoirs as proposed in \cite{vanicat2017exact}. 
This particle system is defined on a generic connected and undirected graph. At each site 
the maximal number of allowed particles {\color{blue}(also called \textit{maximal occupancy})} is $\nu\in \mathbb{N}$. These particles can be of $N-1$ types (or species). Two sites do interact when there is an edge between them and this interaction consists in swapping the occupations of these two sites (for a precise definition of the process see Section~\ref{sec1}). 
Furthermore, reservoirs are attached at each site of the graph, creating and removing particle. The process becomes integrable when the graph is a finite chain with two reservoirs attached to the first and the last sites respectively and the maximal number of allowed particles is fixed to be $\nu=1$,  see \cite{vanicat2017exact}. {\color{blue}As we will see, this choice of the boundary reservoirs is further motivated by the fact that they allow to determine an absorbing dual process, regardless of the geometry on which the process is defined and  the maximal number of particle hosted at each site ($\nu$ can any natural number)}.\\
{\color{blue}In this paper we concentrate our effort in studying the boundary driven symmetric multispecies model. In literature, also asymmetric multispecies  models have been introduced \cite{schutz2017kardar} and integrability results have been proved both with \cite{crampe2016integrable, finn2018matrix} and without boundary driving \cite{mallick1,mallick2}. Moreover, different transition mechanisms (beyond the diffusive one) have been recently studied \cite{chatterjee2023multi}. Finally, the presence of different types of particles have also been modelled via a "multi-layer" dynamics (see \cite{redig2022ergodic,floreani2022switching}). As far as we know, in this last context integrability does not hold.} 
%{\color{red}Explain the model here.... $N$-species (or $N-1$ plus empty?). What is the graph? say that it becomes integrable for the line }
%We refer to the model as the {\em  multispecies stirring process with open boundaries}.
%

@@

{\color{blue} Here $\mu(\bm{\tau})$ is the  non-equilibrium steady probability of having a configuration $\bm{\tau}$, that can be composed by particles and holes as well. However, from Corollary~\ref{Corolollary-ABS_Corr-abstract}, we only have an explicit formula for the expectations of the indicator of particle configurations (indeed $a_{1},\ldots,a_{m}$ are different from $N$). Therefore, to write the probability mass function $\mu$ for occupations containing holes we have to use the conservation of probability to get rid of the indicators of these holes. Then, equation \eqref{implicit-mass-function} is derived (for details see below the proof of Corollary~\ref{Corollary-NESS-ABS}).\\
 The expectation on the right-hand-side of \eqref{implicit-mass-function} contains only indicators of particles configurations: indeed in the first product  $\tau_{x}\neq N$ are considered, while in the second product we have $a_{k}\neq N$ for all $k\in\{1,\ldots,N\}$. As a consequence, we can map the coordinates $x_{1},\ldots,x_{m}$ and the types $a_{1},\ldots,a_{m}$ introduced in Corollary~\ref{Corolollary-ABS_Corr-abstract} with, respectively, the coordinates $x$ (where the $\tau_{x}\neq N$) and $q_{1},\ldots,q_{b}$ and with the types $\tau_{x}\neq N$ and $a_{1},\ldots,a_{b}$. For this reason, by knowing the correlations \eqref{exactCorrelations}, the probability mass function $\mu(\bm{\tau})$ given in equation \eqref{implicit-mass-function} is obtained.}\\

@@

 	Here the right-hand-side corresponds to $b=0$ and average can be computed by Corollary~\ref{Corolollary-ABS_Corr-abstract} with $m=2$, $x_{1}=1,\;x_{2}=2$ and $a_{1}=a_{2}=1$.

@@

 		\mu(1,N)=&\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=1\}}\mathbbm{1}_{\{Y_{2}=N\}}\right]=\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=1\}}\left(1-\sum_{a=1}^{N-1}\mathbbm{1}_{\{Y_{2}=a\}}\right)\right]\nonumber

@@

 	Here the first term on the right-hand-side corresponds to $b=0$ in \eqref{implicit-mass-function} and can be computed by Corollary~\ref{Corolollary-ABS_Corr-abstract} with $m=1$, $x_{1}=1$ and $a_{1}=1$. The second term on the right-hand-side corresponds to $b=1$ in \eqref{implicit-mass-function} and can be computed by Corollary~\ref{Corolollary-ABS_Corr-abstract} with $m=2$, $x_{1}=1,\;x_{2}=2$ and $a_{1}=1,\;a_{2}=a$ for all $a=1,\ldots N-1$. 

@@

 		\mu(N,N)=&\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=N\}}\mathbbm{1}_{\{Y_{2}=N\}}\right]=\mathbb{E}\left[\left(1-\sum_{a_{1}=1}^{N-1}\mathbbm{1}_{\{Y_{1}=a_{1}\}}\right)\left(1-\sum_{a_{2}=1}^{N-1}\mathbbm{1}_{\{Y_{2}=a_{2}\}}\right)\right]\nonumber

@@

 		1-\sum_{a_{1}=1}^{N-1}\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=a_{1}\}}\right]-\sum_{a_{2}=1}^{N-1}\mathbb{E}\left[\mathbbm{1}_{\{Y_{2}=a_{2}\}}\right]+\sum_{a_{1},a_{2}=1}^{N-1}\mathbb{E}\left[\mathbbm{1}_{\{Y_{1}=a_{1}\}}\mathbbm{1}_{\{Y_{2}=a_{2}\}}\right]\dt
 	Here the first term on the right-hand-side corresponds to $b=0$ and since $\bm{\tau}=(N,N)$ we obtain $1$. The second and the third term on the right-hand-side correspond to $b=1$. They can be computed by Corollary~\ref{Corolollary-ABS_Corr-abstract} with $m=1$ and by fixing $x_{1}=1$, $a_{1}=1,\ldots,N-1$ and $x_{2}=2$, $a_{2}=1,\ldots,N-1$ respectively. The fourth term on the right-hand-side corresponds to $b=2$ and it can be computed by Corollary~\ref{Corolollary-ABS_Corr-abstract} with $m=2$, $x_{1}=1,\,x_{2}=2$ and $a_{1},a_{2}=1,\ldots,N-1$. 

@@

			\Psi(\bm{\tau})=&\langle \bm{\tau}|S_{1}^{-1}|\Psi^{'}\rangle=\sum_{\bm{\sigma}\in \Omega^{'}}\Psi^{'}(\bm{\sigma})\prod_{x=1}^{L}\left(\delta_{\tau_{x},\sigma_{x}}+(-1)\delta_{\tau_{x},N}\sum_{a=1}^{N-1}\delta_{\sigma_x,a}\right)\nonumber\\

@@

		means that, {given $\bm{\tau}$ from the bra-vector $\langle \bm{\tau}|$ introduced before equation \eqref{Tensor-S1},} for all $b\in \{0,\ldots,L\}$ and for all $1\leq q_{1}<\ldots<q_{b}\leq L$ {at sites $q_{k}$ (for $k=1,\ldots,b$) we have a particle of type $a_{k}$, while at the other sites the occupation $\tau_{x}$ is left unchanged  ($x\notin\{q_{1},\ldots,q_{b}\}$). }  One can check that equation \eqref{implicit-PSI-PSIp} is indeed formula \eqref{implicit-mass-function} of Corollary~\ref{Corollary-NESS-ABS}. 

@@

\section{Extensions beyond the integrable chain}
\label{sec5}
{\color{blue}In Section~\ref{sectionIntegrabiliy} we have shown that, combining duality and integrability of the multispcies stirring process with $\nu=1$ and on the geometry of chain with two boundaries, the non-equilibrium steady state can be found in an explicit form. As soon as $\nu>1$ the integrability property is lost, but duality still holds. Therefore, in this section we first address the  question whether we can use duality to characterize the non-equilibrium steady state of the boundary driven multispecies stirring process on a chain with $\nu>1$ via absorption probabilities. For the sake of simplicity and  for the sake of clarity, we keep the geometry of a chain with two reservoirs attached to the end sites $1$ and $L$ of the chain. However, this question can be extended to the geometry of the graph $G$ introduced in Section~\ref{sec1}.  \\
Inspired by the results for the thermalized version of the symmetric exclusion process reported in \cite{carinci2013duality}, in this section we also investigate its multispecies counterpart. More precisely, we try to answer these questions: \textit{can we define a thermalized version of the boundary driven multispecies stirring process? Does this thermalized process keep the absorbing duality properties as in the single species case?} Again, we restrict our analysis on the chain with two reservoirs at the end sites. \\ Finally we add a reaction mechanism of transition to the boundary driven multispeices stirring process. The aim is to answer the following question: \textit{Can we generalize to the case with $\nu>1$ with arbitrary number of species the boundary driven model introduced in \cite{casini2022uphill}?} Moreover, in \cite{casini2022uphill} the duality for this reaction-diffusion process was proved only for the equilibrium case and the result was achieved by direct computation (without using a $gl(3)$ symmetry). Thus, we also investigate if this last duality result can be extended to a boundary driven process by using an approach similar to the one of Section~\ref{sectionDuality} and based on Lie algebras.\\ The section is organized as follows: }

@@

	\item \textit{Reaction-diffusion process} (Section~\ref{Subsection-RD}): {\color{blue} starting from the usual stirring dynamics (see Section~\ref{sec1}) where particles and holes are exchanged in two nearest neighbour sites (now at a constant rate $\sigma_{11}$),} {\color{blue}two further reaction} dynamics are added. Particle can now change type {\color{blue} with two different mechanism: a \textit{pure mutation} (with rate $\Upsilon$) and a \textit{stirring-mutation} (at rate $\sigma_{12}$), i.e. a transition where particles perform an exchange of their occupancies at nearest neighbour sites but, at the same time, they also mutate their own species.} {\color{blue} This reaction-diffusion} model   is carefully constructed with the goal of having the average densities evolving as a closed system of $N-1$ coupled difference-differential equation, providing a generalization of the model studied in \cite{casini2022uphill}. {\color{blue} Exploiting the $gl(N)$ symmetry of this reaction-diffusion process, absorbing duality} for the boundary driven process is also proved.

@@

{\color{blue} We recall a general fact. Considering a generic $N-1$ species Markov process $(\bm{n}(t))_{t\geq 0}$ (where in the spirit of Section~\ref{sec1}, $n_{A}^{x}$ denotes the number of particles or the number of holes at site $x$), we define the average density occupation of species $a\in\{1,\ldots,N-1\}$ at site $x$ and at time $t$ as
